---
title: "RMS Titanic"
author: "Cooper Penkava"
date: 2025-02-03
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)


```

**Observations**:

Variables:

-   Class <chr> - refers to the class of ticket the type of passenger had (1st, 2nd, ...)

-   Sex <chr> - refers to the sex of the type of passenger (male, female, ...)

-   Age <chr> - refers to the age range of the type of passenger (child, adult)

-   Survived <chr> - refers to the status of the passenger's life after rescue operations (No, Yes)

-   n <dbl> - refers to the amount of passengers at a certain intersection of the above variables

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Wikipedia lists that there were 2224 people estimated to be on board, but this database accounts for only 2201 people.
-   Are there any differences?
    -   There is a difference of 23 people. This number is not insignificant in the context of human lives.
-   If yes, what might account for those differences?
    -   Lack of contact after the incident - people that simply were unable to be tracked after the disaster may not have been counted in the dataset
    -   people who didn't buy tickets - anyone who did not buy a ticket and/or snuck on may have left evidence of their being there, which wikipedia may account for, but did not leave data to track them with

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   In all passenger categories, more female passengers survived than male passengers - this may imply something about the "women and children first" mentality being employed.
-   However, an extreme amount more of male crew members were saved - I predict that this absolute disproportionate-ness is due to the make up of the data itself, that there were just more male crew members to begin with, but a larger look at the data would need to confirm these thoughts.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Write your observations here.
-   Is there anything *fishy* going on in your plot?
    -   This plot seems to imply that 100% of all passengers in 1st and 2nd class, both male and female survived the tragedy
        -   this should be explored further with more figures and data exploration
    -   It also shows a striking difference from the pure totals graph from before - in the classes that aren't just both 100%, female passengers are at a much higher proportion than men, whereas the former figure showed that just barely more women total were saved in 3rd class and way fewer women were saved in the crew class
        -   this once again goes to show that we need to explore the makeup of the data - this could be explained also by there being much, much more male crew members, and more male 3rd class passengers than females

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Age)
```

**Observations**:

-   (Write your observations here.)
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   This helps us to realize something about how things are plotted in R - before, it was actually plotting the adult and child bars on top of each other, which meant that we were only seeing the overall shape of the larger bar (the child bar)
    -   I think it could be interesting to get a sense of the proportion of each class that is made up of male and female though, like with a pie chart or something.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
