---
title: "Aluminum Data"
author: "Cooper Penkava"
date: 2025-02-16
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

df_stang_long \<-

df_stang %\>%

pivot_longer(

names_to = c("var", "angle"),

names_sep = "\_",

values_to = "val",

starts_with("E") \| starts_with("nu")

) %\>%

mutate(angle = as.integer(angle)) %\>%

pivot_wider(

names_from = var,

values_from = val

)

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>% 
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    values_to = "val",
    starts_with("E") | starts_with("nu")
  ) %>% 
  mutate(angle = as.integer(angle))

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(27, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
##
glimpse(df_stang)
summary(df_stang)
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   no - each different sample has a different value for each of the variables - there is variance, which can come from a multitude of sources all along the manufacturing process of these metal samples
-   How many aluminum alloys are in this dataset? How do you know?
    -   1 - summary lists the alloy column as length 9, but each of those is the same type, as seen when using the glimpse function
-   What angles were tested?
    -   0, 45, and 90, most likely in degrees
-   What thicknesses were tested?
    -   thicknesses including 0.022, 0.032, 0.064, and 0.081
-   Does thickness impact the modulus of elasticity, E? Materials science says it shouldn't, but what does the data say?
    -   let's make a visualization for this one.

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q2 here
df_stang_long %>% 
  ggplot(aes(thick, E)) +
  geom_point()
```

**Observations**:

-   There is not a clear pattern here - I would say the data visualized in this way does not show a trend between sample thickness and modulus of elasticity, E. All the points are around 10,000.
-   It is interesting to note the weird outlier with an E of 0 - this should be explored further
    -   I also wonder why 0.081 thickness deviate a little more from the line than the other three groups - is this a coincidence or is it related to the outier, or is it something unrelated entirely?

### **q4** Consider the following statement:

> "A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

first of all, bruh how did i come up with the same question :( and then you went and made a better graph

-   Does this graph support or contradict the claim above?
    -   I would argue it tends to support it if we make the assumption that the outlier should be thrown out - the first three categories are all basically completely on top each other and the fourth is extremely close.
    -   it would be nice to have a more zoomed in version of this graph after disregarding the outlier just to make sure there are no funny shapes
-   Is this evidence *conclusive* one way or another?
    -   absolutely not - there's only a couple data points here, and one of them is a major outlier and the thickest category is ever so slightly off. I would not trust my bridges, skyscrapers, heavy machinery, and aerospace craft on this evidence.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
