---
title: "Gapminder"
author: "Cooper Penkava"
date: 2025-03-02
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

Looks like these are the following variables:

-   Country

-   Continent

-   Year

-   Life Expectancy

-   Population

-   GDP per Capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>% 
  pull(year) %>% 
  max()
year_min <- 
  gapminder %>% 
  pull(year) %>% 
  min()
year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
first_year <-
  gapminder %>%
  filter(year == year_min)
first_year %>% 
  ggplot(aes(gdpPercap, country)) +
  geom_col() +
  scale_y_discrete(guide = guide_axis(n.dodge=2))

first_year %>% 
  ggplot(aes(gdpPercap)) +
  geom_boxplot()
  
```

**Observations**:

-   Kuwait (I think) has a really high gdp per capita. I wonder why they produce so much money compared to the size of their population?
-   there are a lot more countries that I realized... I thought it was like 100, but I looked it up and there are 193
    -   This bears the question - what about countries that come and go? I don't see the USSR and I do see the Czech republic, both of which contradict the state of the world in the 1950s. At what point do you contradict reality to make your data tidier?
-   I created a second plot to find outliers - a box plot identifies quite a few outliers

**Difficulties & Approaches**:

-   Initially did a basic bar chart but you absolutely couldn't read any country names
-   thought about it, couldn't think of another way to plot qualitative and quantitative data
-   remembered that you plot a bar chart sideways, so i tried that
-   this was only slightly better
-   also, one country (can't read it) is wayyyyyyyy higher than every other one
-   changing the width did not help
-   tried scale x discrete and scale y discrete
-   scale y discrete makes it at least readable

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
cutoff <-
  first_year %>%
  pull(gdpPercap) %>% 
  quantile(prob = .75)
outlier_data <-
  first_year %>% 
  filter(gdpPercap > cutoff)
outlier_countries <-
  outlier_data %>% 
  select(country)
outlier_countries
```

**Observations**:

-   Identify the outlier countries from q2
    -   If we are considering all of the points that are outside the box plot's whiskers to be outliers (as maybe implied by the wording of "possibly many more") then the following are outlier countries.

        |                |     |     |     |     |
        |:---------------|-----|-----|-----|-----|
        | Argentina      |     |     |     |     |
        | Australia      |     |     |     |     |
        | Austria        |     |     |     |     |
        | Bahrain        |     |     |     |     |
        | Belgium        |     |     |     |     |
        | Canada         |     |     |     |     |
        | Chile          |     |     |     |     |
        | Cuba           |     |     |     |     |
        | Czech Republic |     |     |     |     |
        | Denmark        |     |     |     |     |

        |         |     |     |     |     |
        |:--------|-----|-----|-----|-----|
        | Finland |     |     |     |     |
        | France  |     |     |     |     |
        | Gabon   |     |     |     |     |
        | Germany |     |     |     |     |
        | Hungary |     |     |     |     |
        | Iceland |     |     |     |     |
        | Iraq    |     |     |     |     |
        | Ireland |     |     |     |     |
        | Israel  |     |     |     |     |
        | Italy   |     |     |     |     |

|                 |     |     |     |     |
|:----------------|-----|-----|-----|-----|
| Kuwait          |     |     |     |     |
| Lebanon         |     |     |     |     |
| Netherlands     |     |     |     |     |
| New Zealand     |     |     |     |     |
| Norway          |     |     |     |     |
| Poland          |     |     |     |     |
| Saudi Arabia    |     |     |     |     |
| Slovak Republic |     |     |     |     |
| Slovenia        |     |     |     |     |
| South Africa    |     |     |     |     |

|                |     |     |     |     |
|:---------------|-----|-----|-----|-----|
| Sweden         |     |     |     |     |
| Switzerland    |     |     |     |     |
| United Kingdom |     |     |     |     |
| United States  |     |     |     |     |
| Uruguay        |     |     |     |     |
| Venezuela      |     |     |     |     |

-   But to be honest, that's a lot of outliers, so I think I will stick to the top three - the one huge one and the two in the sort up second uppermost grouping

```{r q3-task-redone}
## TASK: Identify the outliers from q2
cutoff <-
  first_year %>%
  pull(gdpPercap) %>% 
  quantile(prob = 190/193) ##top 3 -> (193-3)/193
outlier_data <-
  first_year %>% 
  filter(gdpPercap > cutoff)
outlier_countries <-
  outlier_data %>% 
  select(country)
outlier_countries
```

-   Now, our outliers become Kuwait, Switzerland, and the United States

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

first_year <-
  gapminder %>%
  filter(year == year_min)
first_year %>% 
  ggplot(aes(gdpPercap, country)) +
  geom_col() +
  geom_col(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States")),
    aes(fill = country),
    width = 1
  )+
  scale_y_discrete(guide = guide_axis(n.dodge=2))
```

**Observations**:

-   This makes the top 3 outliers much clearer - before I needed to make the box plot to figure out those top three that the question not so subtly hinted at.
-   I still haven't for the life of me figured out the hella countries problem, but the fill at least lets you know the outlier countries.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(gdpPercap = median(gdpPercap)) %>% 
  ggplot(aes(year, gdpPercap)) +
  geom_line(aes(color = continent))
```

-   Is the world at large becoming more profitable over time? Is the change in GDP equitable between nations?
    -   This graph shows that, over time, the median GDP of each continent has tended to increase, which some fluctuation. The increase is extremely different from continent to continent. Africa is just barely higher at their final value than at their initial value, but Europe and Oceania have increased drastically.

```{r q5-task2}
## TASK: Your second graph
first_and_last_year <-
  gapminder %>%
  filter(year == year_min | year_max)

first_and_last_year %>% 
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot(
    data = . %>% filter(year == year_min),
    aes(color = "Beginning of Data Collection")
  ) +
  geom_boxplot(
    data = . %>% filter(year == year_max),
    aes(color = "End of Data Collection")
  )
```

-   Each of these regions is extremely diverse from country to country - think about how different say the United States are from the countries in South America. How does the variance within continents differ? Is this different at the beginning of when this data was collected compared to the end?
    -   Yes, each of them are very differently varied - but we need to define what we think about when regarding variance - for example, when looking at the final year of the study, Africa has the most outliers, meaning the most points deviating from the IQR, yet it has seeming the second shortest range of data.
    -   From the beginning of the data collection to the end, each of the continents began to capture a larger range of GDPs per capita
    -   Asia went from a very small range to a very large range, a particularly striking change

```{r q5-task3}
## TASK: Your third graph
## graph before identifying outliers
first_year <-
  gapminder %>%
  filter(year == year_min)
first_year %>% 
  ggplot(aes(lifeExp, country)) +
  geom_col() +
  scale_y_discrete(guide = guide_axis(n.dodge=2))

## boxplot to help identify outliers
first_year %>% 
  ggplot(aes(lifeExp)) +
  geom_boxplot()

##identify highest and lowest
cutoff <-
  first_year %>%
  pull(lifeExp) %>% 
  quantile(prob = 192/193) ##top 1 -> (193-1)/193
outlier_data <-
  first_year %>% 
  filter(lifeExp > cutoff)
highest <-
  outlier_data %>% 
  select(country)

cutoff <-
  first_year %>%
  pull(lifeExp) %>% 
  quantile(prob = 1/193) ##bottom 1 -> 1/193
outlier_data <-
  first_year %>% 
  filter(lifeExp < cutoff)
lowest <-
  outlier_data %>% 
  select(country)

highest
lowest

##graph with highest and lowest identified
first_year <-
  gapminder %>%
  filter(year == year_min)
first_year %>% 
  ggplot(aes(lifeExp, country)) +
  geom_col() +
  geom_col(
    data = . %>% filter(country %in% c("Norway", "Afghanistan")),
    aes(fill = c("Lowest", "Highest")),
    width = 1
  )+
  scale_y_discrete(guide = guide_axis(n.dodge=2))
```

-   If we do similar things to the data but focus on life expectancy, what do we see?

    -   Using the same graph but focusing on life expectancy doesn't show a whole lot of interesting anything patterns, but we do see a wide range of life expectancy throughout all the countries.

    -   Looking at the box and whisker plot, we see that there are actually no outliers! Because of this, I think it instead would be interesting to highlight the greatest and smallest life expectancies, as it is somewhat hard to capture that visually with so many bars.

        -   Norway had the highest and Afghanistan had the lowest in the first year that this data was recorded
